Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-11-18 demystify.tex
updates

\subsection{Why CS depts do not waive the application fee?}

@@

In gneral, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admitted to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a prof’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their ``level of excitement'' being in the group.
\end{commentbox}

@@

However, this is not why we ``ghost'' you and provide no response (\autoref{sec:interpreting-response}).  It is because we want to avoid misunderstandings.  A response (\autoref{sec:interpreting-response}), no matter how clear we think it is, might be misinterpreted by the student.  For example, if we say we are not taking students but still (strongly) encourage to apply, then the student might take it as a positive sign of admission. Or if we say we are not interested, the student might take it as a personal rejection and not apply at all.  So it is not because we are too busy to reply, but because we do not want misunderstandings and have to deal with them later.

@@

  \item[Adcom \autoref{sec:evalapps}] Admission Committee. The group of faculty members who review applications and make admission decisions.

  \item [Adcom members \autoref{sec:evalapps}] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.

  \item[Adcom chair \autoref{sec:evalapps}] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).

@@

  \item [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.
  \item [Diversity Statement] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but some universities also require it for graduate admissions. Note that you do not need to be from an underrepresented group to write a diversity statement.

@@

\paragraph{Fee Waiving} Some schools might offer application fee waivers for domestic students.  You should check with the school you're applying to.

>>>>

\subsection{Why CS depts do not waive the application fee?}\label{sec:fee-waive}

@@

In general, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

\subsection{Diversity Statement}\label{sec:diversity-statement}

In recent years many universities start requiring a statement about diversity, equity, and inclusion (DEI) as part of the application (also see \autoref{sec:urm}). This has caused confusion among many applicants, especially international students. They often wonder what to write about, especially if they are not from a minority group, have not experienced anything related the topic, or even not sure what it is about.

A DEI statement is about showing your \emph{awareness of DEI}, your understanding of the challenges others face, and actions you've taken to contribute. Customize what you write to your own background and experiences.  Here are some tips:

\begin{itemize}
\item \textbf{You don't have to be from a minority group.} You just need to understand what DEI is, e.g., be aware of challenges faced by minority groups in studying CS, STEM fields, academia, workplace. Show that you understand the importance of DEI and how it can benefit the CS field and society.

\item \textbf{Show your awareness, customized to your own experiences}. For example, if you come from a different culture or country, talk about DEI here, e.g., many countries completely lack diversity in CS programs, talk about it, what does minority group face in your country?  How do you think DEI can benefit?

\item \textbf{Highlight your contributions}.  You do not need to be in any group or organization specific to DEI.  You can talk about your contributions to your community, e.g., mentoring students, organizing events, or even just helping people in needs. If you don't believe you have done anything much, then talk about what you plan to do.

\item \textbf{Customized it to your application} Can your PhD study or research contribute to DEI?  Can you help or mentor students from underrepresented groups from your country?  These are very doable and can be very convincing.

\begin{commentbox}[Vu:]
One common pitfall is to write about how you are from a poor family, a small town, or a developing country.  These are not related to DEI and raise concerns about you not understanding or even bother to do some basic research about the topic.   I have seen people consider themselves as minority because they are Asian or international students studying in the US (have you seen how many Asian or international students are in CS programs in the US?).\\

Moreover, do not make up stories or exaggerate your experiences.  Just show your awareness and understanding of the topic and what you did or how you plan to contribute to it (e.g., mentoring students, helping others).  One of the best DEI essay I have read start the acknowledgement that the applicant is not a minority, but a white male in a STEM field (i.e., that make them the majority), and that they do not have much experience in DEI, but show that they are aware of the challenges faced by minority groups and how they can contribute to it.
\end{commentbox}

@@

\begin{commentbox}[Hung:]
  I always encourage the students I admit to talk with my students and the students of other faculty in other schools who admitted them. You will unlikely hear straight-out complaints from current students in a prof’s group. But sometimes what is important are things that they (current students) don’t tell you. Pay attention to their ``level of excitement'' being in the group.
\end{commentbox}

@@

However, this is \emph{not} why we ``ghost'' you and provide no response (\autoref{sec:interpreting-response}).  It is because we want to avoid misunderstandings.  A response (\autoref{sec:interpreting-response}), no matter how clear we think it is, might be misinterpreted by the student.  For example, if we say we are not taking students but still (strongly) encourage to apply, then the student might take it as a positive sign of admission. Or if we say we are not interested, the student might take it as a personal rejection and not apply at all.  So it is not because we are too busy to reply, but because we do not want misunderstandings and have to deal with them later.

@@

  \item[Adcom~\autoref{sec:evalapps}] Admission Committee. The group of faculty members who review applications and make admission decisions.

  \item [Adcom members~\autoref{sec:evalapps}] Faculty members who are part of the admission committee. These are the people who review applications and make admission decisions.

  \item[Adcom chair~\autoref{sec:evalapps}] The faculty member who leads the admission committee. This person is often not involved in individual admission decisions but oversees the process (e.g., resolve disputes, ensure fairness).

@@

  \item [Deferred Admission] An option allowing admitted students to postpone their start date, typically by one year.

  \item [Diversity Statement~\autoref{sec:diversity-statement}] A document sometimes required in applications where applicants describe their contributions to diversity and inclusion, or how their background adds to the diversity of the academic community. This document is often required for faculty positions, but more universities are requiring it for graduate admissions. Note that you do not need to be from an underrepresented group to write a diversity statement.

@@

\paragraph{Fee Waiving~\autoref{sec:fee-waive}} Some schools might offer application fee waivers for domestic students.  You should check with the school you're applying to.
